---
title: "Final project EDA"
author: "Kerinna Good and Josh Wagner"
format: pdf
---

## Read in data

``` {r, message = F, warning = F}
library(tidyverse)
library(tidymodels)
library(survival)

acquisitions <- read_csv("data/acquisitions.csv")
all_cetaceans <- read_csv("data/allCetaceanData.csv")
survival_rates <- read_csv("data/survivalRates.csv")
```

Data from [The Pudding](https://data.world/the-pudding/cetaceans). 

```{r}
all_cetaceans2 <- all_cetaceans %>%
  rename(transferdate = transferDate, 
         transfertype = transfer, 
         currentlocation = currently) %>%
  mutate(statusDate2 = ifelse(status == "Alive", "2017-05-07", 
                              all_cetaceans$statusDate)) %>%
  transform(statusyear = as.numeric(substr(statusDate2, 1, 4))) %>%
  mutate(age = as.numeric(statusyear - birthYear),
         originyear = as.numeric(substr(originDate, 1, 4)), 
         captivity = (statusyear - originyear) / age,
         status2 = case_when(status %in% c("Miscarriage", "Escaped (Found Dead)",
                                          "Died", "Stillbirth") ~ 1, 
                             status %in% c("Escaped", "Released", "Released?",
                                           "Unknown", "Alive") ~ 0), 
         acquisition2 = case_when(acquisition %in% c("Born", "Miscarriage", 
                                                     "Stillbirth") ~ "Born",
                                  acquisition %in% c("Capture", "Rescue") ~ 
                                    "Wild", 
                                  acquisition %in% c("Unknown") ~ "Unknown"),
         species2 = ifelse(species == "Bottlenose", "Bottlenose", "Other"), 
         error1 = ((originyear - birthYear) < 0), 
         error2 = ((statusyear - birthYear) < 0),
         transfer = ifelse(originLocation != currentlocation, 1, 0)) %>%
  filter(error1 == FALSE, error2 == FALSE) %>%
  subset(select = -c(id, name, mother, father, region, COD, ...1, entryDate, 
                     notes, statusDate, transfers, error1, error2, status, 
                     acquisition, species))

all_cetaceans2 <- all_cetaceans2[, c(16, 1, 2, 15, 3, 4, 12, 5, 6, 17, 8, 14, 9, 
                                     10, 11, 13)]
```

## Data dictionary for `cetacean_deaths` data set

| Variable        | Class     | Description                                                      	|
|---------------- |---------- |------------------------------------------------------------------	|
| species         | character | Species of animal (whale or dolphin)                             	|
| sex             | character | Sex of animal                                                    	|
| accuracy        | character | Accuracy of an animals' birthdate                               	|
| acquisition     | character | Method through which an animal was brought into captivity       	|
| birthYear       | integer   | Year of birth                                                   	|
| origindate      | integer   | Date that an animal was born into captivity, rescued, or captured |
| originyear      | integer   | Year that an animal was born into captivity, rescued, or captured |
| originlocation	| character	| Location that an animal originated from                          	|
| transfers       | character	| List of facilities an animal was transferred between             	|
| transferdate    | integer   | Date of transfer into the US; NA for animals born in the US      	|
| transfertype    | character | Whether an animal was transferred within the US or outside the US |
| currentlocation | character	| Location of animal at date of recorded status                    	|
| status          | character | Whether an animal was dead or alive at date of recorded status    |
| status2         | binary    | Binary indicator for status (dead or alive)                       |
| statusDate2     | integer   | Date of recorded status                                          	|
| statusyear      | integer   | Year of recorded status                                         	|
| age             | integer   | Age of animal at date of recorded status                          |
| captivity       | integer   | Percentage of animal's life lived in captivity                    |



```{r}
 coxph(formula = Surv(birthYear, statusyear, status2) ~ species2 + sex + accuracy + 
         acquisition2 + captivity + transfertype, data = all_cetaceans2)
#coxph(formula = Surv(birthYear, statusyear, status2) ~ species, data = all_cetaceans2)
```


random change
