---
title: "Final project EDA"
author: "Kerinna Good and Josh Wagner"
format: pdf
---

## Read in data

``` {r, message = F, warning = F}
library(tidyverse)
library(tidymodels)
library(survival)
library(survminer)
library(ggplot2)
library(ggfortify)

all_cetaceans <- read_csv("data/allCetaceanData.csv")
```

Data from [The Pudding](https://data.world/the-pudding/cetaceans). 

```{r, message = F, warning = F, echo = F}
all_cetaceans$statusDate <- as.character(all_cetaceans$statusDate)

all_cetaceans2 <- all_cetaceans %>%
  rename(foreigntransfer = transfer, 
         currentlocation = currently) %>%
  mutate(statusDate2 = ifelse(status == "Alive", "2017-05-07", 
                              all_cetaceans$statusDate)) %>%
  transform(statusyear = as.numeric(substr(statusDate2, 1, 4))) %>%
  mutate(age = as.numeric(statusyear - birthYear),
         originyear = as.numeric(substr(originDate, 1, 4)), 
         captivity = (statusyear - originyear) / age,
         status2 = case_when(status %in% c("Miscarriage", "Escaped (Found Dead)",
                                          "Died", "Stillbirth") ~ 1, 
                             status %in% c("Escaped", "Released", "Released?",
                                           "Unknown", "Alive") ~ 0), 
         acquisition2 = case_when(acquisition %in% c("Born", "Miscarriage", 
                                                     "Stillbirth") ~ "Born",
                                  acquisition %in% c("Capture", "Rescue") ~ 
                                    "Wild", 
                                  acquisition %in% c("Unknown") ~ "Unknown"),
         species2 = ifelse(species == "Bottlenose", "Bottlenose", "Other"), 
         error1 = ((originyear - birthYear) < 0), 
         error2 = ((statusyear - birthYear) < 0),
         transfer = ifelse(originLocation != currentlocation, 1, 0),
         currentlocation2 = case_when(currentlocation == "Unknown" ~ "Unknown", 
                                      currentlocation == "U.S. Navy" ~ "U.S. Navy",
                                      currentlocation == "SeaWorld San Diego" ~ "SeaWorld San Diego", 
                                      currentlocation == "SeaWorld Orlando" ~ "SeaWorld Orlando",
                                      currentlocation %in% c("SeaWorld San Antonio", "Sea World San Antonio") ~ 
                                        "SeaWorld San Antonio", 
                                      currentlocation == "Dolphin Research Center" ~ "Dolphin Research Center",
                                      currentlocation == "Discovery Cove" ~ "Discovery Cove",
                                      currentlocation %in% c("Marineland of Florida", "Marineland Florida") ~
                                                               "Marineland Florida",
                                      currentlocation == "Miami Seaquarium" ~ "Miami Seaquarium",
                                      currentlocation == "Marine Life Oceanarium" ~"Marine Life Oceanarium",
                                      .default = "aOther")) %>%
  filter(error1 == FALSE, error2 == FALSE) %>%
  subset(select = -c(id, name, mother, father, region, COD, ...1, entryDate, 
                     notes, statusDate, transfers, error1, error2, status, 
                     acquisition, species))

all_cetaceans2 <- all_cetaceans2[, c(16, 1, 2, 15, 3, 4, 12, 5, 6, 18, 17, 8, 14, 9, 
                                     10, 11, 13)]
```

```{r}

all_cetaceans2 %>%
  count(captivity)

all_cetaceans2$birthDate <- as.Date(paste(all_cetaceans2$birthYear, "-07", "-01", sep=""))
all_cetaceans2$birthDate <- as.character(all_cetaceans2$birthDate) 
all_cetaceans2$originDate <- as.character(all_cetaceans2$originDate)

all_cetaceans2 <- all_cetaceans2 %>%
  mutate(birthDate2 = ifelse(acquisition2 == "Wild", birthDate, originDate))

all_cetaceans2$birthDate2 <- as.Date(all_cetaceans2$birthDate2)
all_cetaceans2$originDate <- as.Date(all_cetaceans2$originDate)

all_cetaceans2 <- all_cetaceans2 %>%
  mutate(age2 = difftime(statusDate2, birthDate2, units = "day"),
         captivity = difftime(statusDate2, originDate, units = "day"), 
         captivity_prop = (as.numeric(captivity) / as.numeric(age2)))

#all_cetaceans2 <- all_cetaceans2[, c(16, 1, 2, 15, 3, 4, 12, 5, 6, 17, 8, 14, 9, 
#                                     10, 11, 13)]

all_cetaceans %>%
  count(COD) %>%
  arrange(desc(n))

all_cetaceans %>%
  count(currently) %>%
  arrange(desc(n))
```

## Data dictionary for `all_cetaceans2` data set

| Variable        | Class     | Description                                                      	|
|---------------- |---------- |------------------------------------------------------------------	|
| species         | character | Species of animal (whale or dolphin)                             	|
| sex             | character | Sex of animal                                                    	|
| accuracy        | character | Accuracy of an animal's birth date                               	|
| acquisition2    | character | Method through which an animal was brought into captivity       	|
| birthYear       | integer   | Year of birth                                                   	|
| originDate      | integer   | Date that an animal was born into captivity, rescued, or captured |
| originyear      | integer   | Year that an animal was born into captivity, rescued, or captured |
| originlocation	| character	| Location that an animal originated from                          	|
| currentlocation | character | Location of animal at date of recorded status                     |
| transfer        | binary  	| Whether current location matches origin location                	|
| foreigntransfer | character | Whether an animal was transferred from outside the US             |
| status2         | binary    | Binary indicator for status (dead or alive)                       |
| statusDate2     | integer   | Date of recorded status                                          	|
| statusyear      | integer   | Year of recorded status                                         	|
| age             | integer   | Age of animal at date of recorded status                          |
| captivity       | integer   | Percentage of animal's life lived in captivity                    |

## Exploratory data analysis

```{r}
model_fit1 <- survfit(Surv(age, status2) ~ species2, 
      data = all_cetaceans2)
autoplot(model_fit1) + 
 labs(x = "\n Survival time (years) ", y = "Survival probabilities \n", 
 title = "Survival time of cetaceans in captivity by species", 
 color = "Species")

model_fit2 <- survfit(Surv(age, status2) ~ sex, 
      data = all_cetaceans2)
autoplot(model_fit2) + 
 labs(x = "\n Survival time (years) ", y = "Survival probabilities \n", 
 title = "Survival time of cetaceans in captivity by sex", 
 color = "Sex")

model_fit3 <- survfit(Surv(age, status2) ~ acquisition2, 
      data = all_cetaceans2)
autoplot(model_fit3) + 
 labs(x = "\n Survival time (years) ", y = "Survival probabilities \n", 
 title = "Survival time of cetaceans in captivity by acquisition method", 
 color = "Acquisition method")

model_fit4 <- survfit(Surv(age, status2) ~ transfer, 
      data = all_cetaceans2)
autoplot(model_fit4) + 
 labs(x = "\n Survival time (years) ", y = "Survival probabilities \n", 
 title = "Survival time of cetaceans in captivity by transfer status", 
 color = "Transfer")

model_fit5 <- survfit(Surv(age, status2) ~ foreigntransfer, 
      data = all_cetaceans2)
autoplot(model_fit5) + 
 labs(x = "\n Survival time (years) ", y = "Survival probabilities \n", 
 title = "Survival time of cetaceans in captivity by transfer type", 
 color = "Transfer type")

model_fit6 <- survfit(Surv(age, status2) ~ currentlocation2, 
      data = all_cetaceans2)
autoplot(model_fit6) + 
 labs(x = "\n Survival time (years) ", y = "Survival probabilities \n", 
 title = "Survival time of cetaceans in captivity by current location")

all_cetaceans2 %>%
  ggplot(mapping = aes(x = captivity)) +
  geom_histogram(binwidth = 0.02) +
  labs(x = "Percentage of life lived in captivity", y = "Count", 
       title = "Most cetaceans live majority of their lives in captivity")

all_cetaceans2 %>%
  subset(select = c(age, status2, species2, sex, acquisition2, captivity,
                    transfer, foreigntransfer)) %>%
  tbl_summary() %>%
  as_gt()
```
Code for survival visualizations from [dkmathstats Website](https://dk81.github.io/dkmathstats_site/rvisual-kaplan-meier.html).


## Model selection

```{r}
model1 <- coxph(formula = Surv(age, status2) ~ species2 + sex + 
         acquisition2 + captivity + transfer + foreigntransfer + currentlocation2, 
         data = all_cetaceans2)

model1

ggsurvplot(survfit(model1), data = all_cetaceans2, color = "#2E9FDF",
           ggtheme = theme_minimal()) +
  labs(title = "Survival rate of cetaceans in captivity", x = "Time (days)")
```
Code for visualization from [Statistical tools for high-throughput data analysis](http://www.sthda.com/english/wiki/cox-proportional-hazards-model)



